\pkg{sparseHessianFD}: An R Package for Estimating Sparse Hessian Matrices

### 2016-12-25 nobuild/jss_revision/response.tex

\usepackage{geometry}
\geometry{margin=1.1in}

\setlength{\parskip}{1ex}
\setlength{\parindent}{0ex}

\begin{document}

{\bfseries
\begin{center}
  Manuscript JSS2717\\
 ``sparseHessianFD: An R Package for Estimating Sparse
 Hessian Matrices''\\
 \vspace{\baselineskip}
 Response to reviewer
\end{center}
}
\vspace{\baselineskip}

I would like to express my sincere appreciation for the time and effort that
was invested in the review.  I am happy to write that I was
able to incorporate a large majority of the suggestions into the the
revised paper and package. For the
others, I hope that you and the editor will accept my
explanations and excuses.  Of course, if there are any strong
objections to any of my responses, I am happy
to consider further changes in advance of publication.

@@

(error messages related to algorithm.sty removed).

I don't remember if there is an easy way to distribute ``special'' sty files,
or if you must force the user to install them. If the sty is not really
critical it might be better to omit it.

\end{revQuote}

\begin{response}
Because \pkg{sparseHessianFD} passes CRAN check, I suspect that the
error is a result of the \pkg{algorithm} package
not being present in the LaTeX installation.  It is a standard
package in the MacTeX distribution, but I cannot speak with authority about other
platforms. It should be easily installed through CTAN.

@@

I would prefer to continue to use the \pkg{algorithm} package, but I
could try to find another way to format the algorithms if there is a
strong objection.
\end{response}

\item \begin{revQuote}
I am not familiar with reference classes, but for most users I think it
should not be important that the package uses them. However, as a user, I
find it disconcerting that sparseHessianFD() is used as a (constructor)
function (eg. p12, and in the example in ?sparseHessianFD) but is not
documented in the help with 'usage' and 'arguments' as is usual for
functions. There seems to be little guidance in the usual R places about how
reference classes should be documented. Perhaps you could seek some guidance
from the community on this point.
\end{revQuote}

\begin{response}
The help page for \pkg{sparseHessianFD} now includes explicit
instructions for the initializer.  It is correct that there is
little guidance in how to document reference classes, and I agree that
there should be a way for the
existing mechanism to include documentation for the initializer to
an RC.  But it's hard to do it using \pkg{roxygen2} in way that also passes
CRAN checks. Therefore, I just ``hard coded'' a description of the initializer
into the details section of the sparseHessianFD-class documentation. I
could not think of a more elegant way of doing it.  
\end{response}

@@

\item \begin{revQuote}
It is pointed out (p2) that requirements are burdensome and emphasized
that the package is not appropriate for all uses. Also, the conditions 1-5
are fairly stringent. The reader is left with the impression that there may
be few applications. I think the value of the package could be promoted
better. Perhaps a small list of different types or applications of
hierarchical problems could be given, or an indication of other classes of
problems where the conditions would apply.

  \end{revQuote}

  \begin{response}
In this original submission I included this section primarily as a way to avoid overselling the
method, but your point that the reader might be scared away by this list of
restrictions is well-taken.  In the revision, I removed the list, and
wove the most important restrictions into the text.  The revision projects a more
positive tone that in the original submission.
\end{response}

@@

\begin{response}
The descriptions of the two sparsity patterns are correct. The
first index in the subscript of $\beta$ refers to a household, and the
second index refers to a covariate. Thus,
$\parD{2}{\beta_{11},\beta_{12}}{}\neq 0$, but
$\parD{2}{\beta_{11},\beta_{21}}{}=0$.  When the second index changes
faster than the first, the pattern is ``block-arrow.''  When the first
index changes faster than the second, the pattern is ``banded.''

\end{response}

\item \begin{revQuote}
Code in the paper prior to table 4 needs to be copied from the
vinettes/sparseHessianFD.Rnw file. For those trying to reproduce results it
would be nice if this where mentioned in the file replication.R.
  \end{revQuote}

  \begin{response}
Assuming that the \proglang{R} package builder works the way I think it does,
the \filename{replication.R} file, and the data for Table 4 (\filename{vignette\_tab4.Rdata}), will be
copied to the \filename{doc\textbackslash} directory of the package
once it is installed.  At the end of the second paragraph of Section
4, I now write that ``code to replicate Table 4 is available as an
online supplement to this paper.'' In the non-JSS vignette that will be part
of the package (and posted on CRAN), this sentence will be changed appropriately.
\end{response}

@@

  \end{revQuote}

\begin{response}

Here I am combining my responses to items \ref{it:complex1} and
\ref{it:complex2} together.

  The complex step function is now implemented in the package, and is
  described as one of two methods of numerical approximation (the
  other being finite differences).  It is 
  correct that ``most R primitives work with complex numbers,''
  but there are many commonly used functions that do not, including 
  \func{log1p}, \func{gamma}, and the probability distribution
  functions.  This limitation, along
  with the requirement that the function be
  holomorphic, are discussed in Section 2.4

@@

In terms of accuracy, the paper now includes comparisons between the true
Hessian and the two \class{sparseHessianFD} methods (finite differences
and complex step). To maintain the focus of the paper on
\pkg{sparseHessianFD}, I am not including an assessment of the accuracy of
\pkg{numDeriv}.  However, for the
timing comparisons in Table 4, I did
change the baseline \pkg{numDeriv} method from \func{hessian} with the
\method{Richardson} method, to \func{jacobian} with both the
\method{simple} and 
\method{complex} methods.

I must admit that I was not familiar with the complex step method
beforehand.  Incorporating it into the package, and writing about it
in the paper, was a valuable learning experience for me. That said,
until now I have \emph{never} come across complex analysis in my own work, and I
may not have done anything with complex numbers at all since my
Algebra II class in 10th grade. Thus, it is highly probable that the parts
of the paper on the complex step method are not precisely correct.
Any constructive criticism or suggestions in that regard would be
warmly welcomed.

@@

  \end{revQuote}

\begin{response}
  I added a \func{set.seed} statement near the top of
  \filename{replication.R}, but it is correct that because only
  times are reported, this does not matter much.
\end{response}

\item \begin{revQuote}
Table 4 and 5. Some OS, processor, and memory details are helpful to put
timing results in context.
  \end{revQuote}

\begin{response}
  This information is now included in Section 4.
\end{response}

@@

(On my laptop the next took about 30 hrs of which 24 was for the last,
N=2500, k=8 comparison.)

(Code and output removed)

  \end{revQuote}

  \begin{response}
Because the \func{numDeriv::jacobian} function runs so much faster
than  \func{numDeriv::hessian}, I was able to increase the size of the
examples in Table 4: $N=(15, 50, 100, 500)$ and $k=(2,5,8)$. Please
excuse me for not going as high as $N=2500$, $k=8$. For the reader
who wants to replicate the results, a run time of more than a day is a
lot to ask.
  \end{response}

\item\begin{revQuote}
 p.16  l. -7 "to to compute" -> "to compute"
  \end{revQuote}

@@

  \begin{response}
    Fixed.  Thanks.
  \end{response}

\item\begin{revQuote}
While that package is useful, and reasonably demonstrated in the paper,
I think it would be nice to expand the paper in some ways that might be
deemed more "original research". Some possibilities are:
\begin{itemize}
 \item Explain and try to assess how much of the speedup is due to simple
sparseness and how much is due to the "sparse patern" (p2) allowing for
perturbing multiple elements together. (I think these are related but
slightly different?)

@@

 \item Try to assess how much of the speedup is due to reduce computing demand
and how much is due to different memory demand. (I had the impression in the
larger problems with numDeriv that my computer started to use swap space,
which resulted in a big slow down.)

 \item Consider implementing a complex step method, and do a comparison.

 \item  Assess the difference when multiple CPUs are used. (run.par == FALSE vs
run.par == TRUE)
\end{itemize}
  \end{revQuote}

  \begin{response}
I appreciate the editor's comment in the decision letter that these suggestions might be
better left for a future paper.  But they are reasonable points that I
would like to address in turn.
    \begin{itemize}
    \item This is discussed in Section 3.1.  In general, it really is all about the sparsity pattern, and
      specifically, the how few partitions are needed to group the
      variables.  The ``simple sparseness'' comes into play because it
      is more likely the number of partitions will be small with the
      the proportion of structural zeros is large. The \emph{arrangement} of the non-zeros is
      related to the ordering of the variables, but the permutation
      handles that during the initialization of the
      \class{sparseHessianFD} argument.  For example, variables
      arranged in a way that the Hessian has a block-arrow pattern are
      permuted to get a Hessian with a banded pattern instead.

### 2016-12-25 nobuild/jss_revision/sparseHessianFD.tex
copied jss article to vignette

The \pkg{sparseHessianFD} package is a tool for estimating
sparse Hessians numerically, using either finite differences or
complex perturbations of
gradients.  Section~\ref{sec:numdiff} will
cover the specifics, but the basic idea is as follows.  Consider a real-valued
function $f(x)$, its gradient $\nabla{f(x)}$, and its Hessian
$\hess{}{f(x)}$, for $x\in\mathbb{R}^M$.  Define the derivative vector
as the transpose of the gradient, and a vector of partial derivatives, so
${\parD{}{}{f(x)}=\nabla f(x)^\top=\left(\parD{}{1}{},\mathellipsis,\parD{}{M}{}\right)}$.   (Throughout the paper, we will try
to reduce notational clutter by referring to the derivative and
Hessian as $\parD{}{}{}$ and $\hess{}{}$, respectively, without the
$f(x)$ symbol).  Let $e_m$ be a vector of zeros, except with a 1 in the $m$th
element, and let $\delta$ be a sufficiently small scalar
constant. A ``finite difference'' linear approximation to the $m$th column of the Hessian is ${\hess{m}{}\approx\left(\nabla f(x+\delta
  e_m)- \nabla f(x)\right)/\delta}$.  Estimating a dense Hessian in this way involves at least $M+1$ calculations of the
gradient: one for the gradient at $x$, and one after perturbing each
of the $M$ elements of $x$, one at a time. If the function is complex
analytic, an alternative approximation that requires $M$ gradient
evaulations is $\hess{m}{}\approx\Imag\left(\nabla f(x+i\delta
  e_m)\right)/\delta$, where $i=\sqrt{-1}$ and $\Imag$ returns the imaginary part of a complex
number \citep{SquireTrapp1998}. This ``complex step'' method can be more accurate than the
finite difference method because there is no subtraction operation
that might generate roundoff errors, and thus $\delta$ can be made
arbitrarily small to avoid truncaton errors \citep{LaiCrassidis2008}. However, it does require that the
implementation of $f(x)$ accept a complex argument, and return a
complex result, even though we are interested in differentiating $f(x)$ only along
the real line.

Regardless of the approximation method used, if the Hessian is sparse, most of the
elements are constrained to zero.  Depending
on the sparsity pattern of the Hessian, those constraints may let us recover
the Hessian with fewer gradient evaluations by perturbing multiple
elements of $x$ together.  For some sparsity patterns, estimating a Hessian in
this way can be profoundly efficient.  In fact, for the hierarchical models that we consider in
this paper, the number of gradient evaluations does not increase with
the number of additional heterogeneous units.

@@

\begin{table}[ptb]\centering
  \begin{tabularx}{1\linewidth}{@{\extracolsep{\fill} }rrr|rrrr|rrrr}
  \toprule
&&&\multicolumn{4}{c|}{\bfseries{Finite differencing}}&\multicolumn{4}{c}{\bfseries{Complex step}}\\
&&&\multicolumn{2}{c}{\pkg{numDeriv}}&\multicolumn{2}{c|}{\pkg{sparseHessianFD}}&\multicolumn{2}{c}{\pkg{numDeriv}}&\multicolumn{2}{c}{\pkg{sparseHessianFD}}\\
$N$&$k$&$M$&mean&sd&mean&sd&mean&sd&mean&sd\\
%%\midrule
  \midrule
15 & 2 & 32 & 12.4 & 1.2 & 2.3 & 0.5 & 12.8 & 1.2 & 2.1 & 0.5 \\ 
  15 & 4 & 64 & 26.2 & 12.4 & 4.0 & 0.9 & 26.1 & 2.0 & 3.8 & 1.1 \\ 
  50 & 2 & 102 & 50.5 & 2.4 & 3.0 & 0.9 & 54.8 & 3.5 & 2.9 & 2.1 \\ 
  15 & 8 & 128 & 50.4 & 2.2 & 8.5 & 12.0 & 55.0 & 2.7 & 7.7 & 1.5 \\ 
  100 & 2 & 202 & 134.1 & 15.7 & 3.7 & 0.5 & 144.0 & 11.6 & 3.3 & 0.6 \\ 
  50 & 4 & 204 & 104.0 & 2.1 & 6.3 & 11.6 & 117.2 & 17.6 & 5.1 & 1.0 \\ 
  100 & 4 & 404 & 284.3 & 34.3 & 8.0 & 10.9 & 322.5 & 40.8 & 8.2 & 11.1 \\ 
  50 & 8 & 408 & 248.5 & 48.7 & 10.9 & 1.4 & 271.7 & 43.7 & 11.5 & 1.4 \\ 
  100 & 8 & 808 & 624.5 & 54.0 & 14.4 & 1.8 & 762.6 & 76.9 & 15.7 & 1.7 \\ 
  500 & 2 & 1002 & 1758.6 & 86.7 & 10.1 & 10.0 & 1990.0 & 90.6 & 9.3 & 9.8 \\ 
  500 & 4 & 2004 & 3832.5 & 84.9 & 18.4 & 2.3 & 4460.7 & 82.8 & 19.4 & 2.5 \\ 
  500 & 8 & 4008 & 8846.5 & 82.1 & 45.2 & 6.2 & 11058.4 & 101.3 & 52.5 & 6.3 \\ 
   \bottomrule

@@

To help us understand just how scalable
\pkg{sparseHessianFD} is, we ran another set of simulations, for the
same hierarchical model, for different values of $N$ and $k$.  We then
recorded the run times for different steps in the sparse Hessian
estimation, across 200 replications.  The steps are summarized in
Table~\ref{tab:timeMeasures}.  The times were generated on an
Apple Mac Pro with a 12-core Intel Xeon E5-2697 processor (2.7 GHz)
with 64 GB of RAM.

>>>>

The \pkg{sparseHessianFD} package is a tool for estimating
sparse Hessians numerically, using either finite differences or
complex perturbations of
gradients.  Section~\ref{sec:numdiff} will
cover the specifics, but the basic idea is as follows.  Consider a real-valued
function $f(x)$, its gradient $\nabla{f(x)}$, and its Hessian
$\hess{}{f(x)}$, for $x\in\mathbb{R}^M$.  Define the derivative vector
as the transpose of the gradient, and a vector of partial derivatives, so
${\parD{}{}{f(x)}=\nabla f(x)^\top=\left(\parD{}{1}{},\mathellipsis,\parD{}{M}{}\right)}$.   (Throughout the paper, we will try
to reduce notational clutter by referring to the derivative and
Hessian as $\parD{}{}{}$ and $\hess{}{}$, respectively, without the
$f(x)$ symbol).  Let $e_m$ be a vector of zeros, except with a 1 in the $m$th
element, and let $\delta$ be a sufficiently small scalar
constant. A ``finite difference'' linear approximation to the $m$th column of the Hessian is ${\hess{m}{}\approx\left(\nabla f(x+\delta
  e_m)- \nabla f(x)\right)/\delta}$.  Estimating a dense Hessian in this way involves at least $M+1$ calculations of the
gradient: one for the gradient at $x$, and one after perturbing each
of the $M$ elements of $x$, one at a time. Under certain conditions, a
more accurate approximation is the ``complex step'' method:  $\hess{m}{}\approx\Imag\left(\nabla f(x+i\delta
  e_m)\right)/\delta$, where $i=\sqrt{-1}$ and $\Imag$ returns the
imaginary part of a complex
number \citep{SquireTrapp1998}.  Regardless of the approximation method used, if the Hessian is sparse, most of the
elements are constrained to zero.  Depending
on the sparsity pattern of the Hessian, those constraints may let us recover
the Hessian with fewer gradient evaluations by perturbing multiple
elements of $x$ together.  For some sparsity patterns, estimating a Hessian in
this way can be profoundly efficient.  In fact, for the hierarchical models that we consider in
this paper, the number of gradient evaluations does not increase with
the number of additional heterogeneous units.

@@

\begin{table}[ptb]\centering
  \begin{tabularx}{1\linewidth}{@{\extracolsep{\fill} }rrr|rrrr|rrrr}
  \toprule
&&&\multicolumn{4}{c|}{\bfseries{Finite differencing}}&\multicolumn{4}{c}{\bfseries{Complex step}}\\
&&&\multicolumn{2}{c}{\pkg{numDeriv}}&\multicolumn{2}{c|}{\pkg{sparseHessianFD}}&\multicolumn{2}{c}{\pkg{numDeriv}}&\multicolumn{2}{c}{\pkg{sparseHessianFD}}\\
$N$&$k$&$M$&mean&sd&mean&sd&mean&sd&mean&sd\\
%%\midrule
  \midrule
15 & 2 & 32 & 12.7 & 0.9 & 2.4 & 0.1 & 13.2 & 0.7 & 2.0 & 0.1 \\ 
  15 & 5 & 80 & 32.2 & 0.9 & 5.0 & 0.5 & 36.1 & 10.1 & 5.2 & 0.8 \\ 
  50 & 2 & 102 & 51.3 & 9.8 & 3.0 & 0.4 & 54.5 & 1.0 & 2.7 & 0.6 \\ 
  15 & 8 & 128 & 53.6 & 0.9 & 7.8 & 0.6 & 62.5 & 16.7 & 8.2 & 0.8 \\ 
  100 & 2 & 202 & 131.7 & 16.9 & 3.7 & 0.3 & 147.6 & 24.1 & 3.4 & 0.5 \\ 
  50 & 5 & 255 & 144.0 & 27.1 & 6.6 & 0.6 & 162.8 & 29.2 & 6.7 & 0.7 \\ 
  50 & 8 & 408 & 256.2 & 43.5 & 12.5 & 12.9 & 291.1 & 39.0 & 11.7 & 0.7 \\ 
  100 & 5 & 505 & 495.7 & 129.5 & 9.8 & 3.1 & 559.8 & 125.9 & 11.8 & 10.4 \\ 
  100 & 8 & 808 & 850.0 & 171.4 & 20.9 & 20.4 & 999.2 & 217.5 & 23.6 & 17.3 \\ 
  500 & 2 & 1002 & 1953.4 & 321.7 & 10.8 & 3.7 & 2156.2 & 339.7 & 9.5 & 3.4 \\ 
  500 & 5 & 2505 & 5444.0 & 951.0 & 27.1 & 10.6 & 6159.3 & 1009.2 & 26.5 & 4.6 \\ 
  500 & 8 & 4008 & 9533.1 & 190.9 & 54.5 & 12.8 & 14059.4 & 3797.9 & 62.5 & 21.8 \\ 
   \bottomrule

@@

To help us understand just how scalable
\pkg{sparseHessianFD} is, we ran another set of simulations, for the
same hierarchical model, for different values of $N$ and $k$.  We then
recorded the run times for different steps in the sparse Hessian
estimation, across 500 replications.  The steps are summarized in
Table~\ref{tab:timeMeasures}.  The times were generated on an
Apple Mac Pro with a 12-core Intel Xeon E5-2697 processor (2.7 GHz)
with 64 GB of RAM.
